---
title: "Survival_analysis_project"
output:
  html_document:
    toc : false 
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,        # show code (or set to FALSE if you want to hide it)
  message = FALSE,    # hide all package load messages
  warning = FALSE,    # hide all warnings
  error = FALSE    # hide errors in knitting
)

```

```{r}
#install.packages("patchwork")
library(patchwork)
#install.packages("ggcorrplot")
library(ggcorrplot)
#install.packages("dplyr")
library(dplyr)


# Load packages
library(survival)
library(knitr)
library(ggplot2)
library(skimr)
library(corrplot)
library(survminer)
library(dplyr)
library(MASS)
library(tidyr)
```

```{r}
#download the files :

df <- read.csv("C:/Users/tamar/OneDrive/uni/4th/survival_analysis/SEER Breast Cancer Dataset .csv")
#colnames(df)


#Age - numaric
#Race - cat - Black/ other/ white
#Marital Status - cat -  Divorced / married /single / widow/ seprated
#T Stage - cat -  T1 /T2/T3/T4
#N Stage - cat - N1/N2/N3
#6th Stage - cat - IIA / IIB/ IIIA / IIIB / IIIC
#Grade - cat - Moderately differentiated; Grade II / Poorly differentiated; Grade III / Well differentiated; Grade I / Undifferentiated; anaplastic; Grade IV
#A Stage -cat - Distant / Ragional
#Tumor Size - numric
#Estrogen Status - cat - "Positive" "Negative"
#Progesterone Status - cat - "Positive" "Negative"
#Regional Node Examined - numaric
#Reginol Node Positive - numric
#Survival Months -  numric
#Status - cat -  "Alive" "Dead" 




df_seer_clean <- df %>%
  mutate(
    # Numeric variables
    Age                     = as.numeric(Age),
    Tumor.Size              = as.numeric(Tumor.Size),
    Regional.Node.Examined  = as.numeric(Regional.Node.Examined),
    Reginol.Node.Positive   = as.numeric(Reginol.Node.Positive),
    Survival.Months         = as.numeric(Survival.Months),
    status                  = ifelse(Status == "Dead", 1, 0),  # 1 = dead, 0 = alive
    
    # Categorical variables as regular factors
    Race           = factor(Race),
    Marital.Status = factor(Marital.Status),
    T.Stage        = factor(T.Stage),
    N.Stage        = factor(N.Stage),
    X6th.Stage     = factor(X6th.Stage),
    Grade          = factor(Grade),
    A.Stage        = factor(A.Stage),
    Estrogen.Status     = factor(Estrogen.Status),
    Progesterone.Status = factor(Progesterone.Status)
  ) %>%
  
  # reorder columns to put survival variables first
  dplyr::select(
    status,
    Age, Race, Marital.Status,
    T.Stage, N.Stage, X6th.Stage, Grade, A.Stage,
    Tumor.Size,
    Estrogen.Status, Progesterone.Status,
    Regional.Node.Examined, Reginol.Node.Positive,
    everything()
  )
```

```{r}
skim(df_seer_clean)

```


```{r}

#plots for numerics :
numeric_vars <- df_seer_clean[, sapply(df_seer_clean, is.numeric)]


#  Histograms of numeric variables


for (v in names(numeric_vars)) {
  print(
    ggplot(df_seer_clean, aes(x = .data[[v]])) +
      geom_histogram(bins = 30, fill = "lightblue", color = "white") +
      theme_minimal() +
      labs(
        title = paste("Histogram of", v),
        x = v,
        y = "Count"
      )
  )
}



```
```{r}
#plots for cat
library(ggplot2)
library(dplyr)

# Select categorical variables
cat_vars <- df_seer_clean[, sapply(df_seer_clean, is.factor)]

# Loop and plot each categorical variable
for (v in names(cat_vars)) {
  
  plot_obj <- ggplot(df_seer_clean, aes(x = .data[[v]])) +
    geom_bar(fill = "lightgreen", color = "white") +
    geom_text(
      stat = "count",
      aes(label = ..count..),
      vjust = -0.3,
      size = 3
    ) +
    theme_minimal() +
    labs(
      title = paste("Bar Plot of", v),
      x = v,
      y = "Count"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)  # rotate labels for readability
    )
  
  print(plot_obj)
}

```

```{r}

library(corrplot)

# Select numeric variables for correlation
corr_vars <- df_seer_clean[, c(
  "Age",
  "Tumor.Size",
  "Regional.Node.Examined",
  "Reginol.Node.Positive",
  "Survival.Months"
  # You can add "status" if you want: "status"
)]

# Compute Spearman correlation matrix
cor_mat <- cor(corr_vars, method = "spearman", use = "complete.obs")
cor_mat


corrplot(
  cor_mat,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  addCoef.col = "black",   # add correlation numbers
  number.cex = 0.8,
  col = colorRampPalette(c("darkblue", "white", "darkred"))(200),
  title = "Correlation Among Numeric Variables in the SEER Breast Cancer Dataset",
  mar = c(0,0,3,0)
)


```

#### The numeric variables in the dataset show weak correlations overall, except for a modest relationship between nodes examined and nodes positive, and a perfect expected identity between time and Survival.Months. There is no harmful multicollinearity, so all numeric predictors are suitable for use in survival models.

-------------
Handling ties
-------------

```{r}
t_event <- df_seer_clean$Survival.Months[df_seer_clean$status == 1]     # Extract survival times for events only.

n_event <- length(t_event)                                              # Total number of observed events.
n_unique_event <- length(unique(t_event))                               # Number of unique event times.
cat("Events:", n_event, "\nUnique event times:", n_unique_event, "\n")
cat("Fraction tied (events):", 1 - n_unique_event/n_event, "\n")
```
```{r}
ties_choice <- "efron"  # Heavy ties in event times (≈84%); Efron is preferred over Breslow and discrete is too heavy.
```

-------------------------------
Handling right-skewed variables
-------------------------------

```{r}
# Creating log(1 + x) columns for right-skewed columns
df_seer_clean$Tumor.Size_log <- log1p(df_seer_clean$Tumor.Size)   
df_seer_clean$Regional.Node.Examined_log <- log1p(df_seer_clean$Regional.Node.Examined)   
df_seer_clean$Reginol.Node.Positive_log <- log1p(df_seer_clean$Reginol.Node.Positive)  

# Checking the new distribution
vars_to_plot <- c(
  "Tumor.Size_log",
  "Regional.Node.Examined_log",
  "Reginol.Node.Positive_log"
)

for (v in vars_to_plot) {
  p <- ggplot(df_seer_clean, aes(x = .data[[v]])) +
    geom_histogram(bins = 30, fill = "lightblue", color = "white") +
    theme_minimal() +
    labs(
      title = paste("Histogram of", v),
      x = v,
      y = "Count"
    )
  print(p)
}

```
After applying the log(1 + x) transformation, the distributions are noticeably less right-skewed and the long tails are compressed. This makes the variables more well-behaved for modeling.



```{r}
# Create a new feature: node_ratio = (number of positive nodes) / (number of examined nodes).

df_seer_clean <- df_seer_clean %>%
  mutate(
    node_ratio = case_when(
      Regional.Node.Examined == 0 & Reginol.Node.Positive == 0 ~ 0,         # examined = 0 and positive = 0  -> define node_ratio = 0
      Regional.Node.Examined == 0 & Reginol.Node.Positive  > 0 ~ NA_real_,  # examined = 0 and positive > 0  -> set node_ratio = NA
      TRUE ~ Reginol.Node.Positive / Regional.Node.Examined
    )
  )

ggplot(df_seer_clean, aes(x = node_ratio)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "white") +
  theme_minimal() +
  labs(title = "Histogram of node_ratio", x = "node_ratio", y = "Count")

```










May, for later on, when running cox, make sure you put this ties option.
coxph(Surv(Survival.Months, status) ~ ..., data = df_seer_clean, ties = ties_choice)

